\FOO: A framework for efficient source code generation\\ of detection algorithms for the Internet of Things

### 2015-04-17 paper-euspn15.tex
final reading before draft release

The ability to detect patterns in its usage is what makes \emph{things}
\emph{smart}. The more patterns it can detect, the more accurate its
operational results will be. However, supporting an adequate set of pattern
detection algorithms imposes significant overhead on resource-constrained
devices such as those in the Internet of Things. Manual fusion of the
algorithms reduces resource consumption by optimizing resource sharing, yet,
proves to be time-consuming, repetitive and error-prone. To address this
problem we propose \FOO, a framework for the generation of efficient
combinations of detection algorithms. It consists of a domain specific language
and a code generator. The language allows formally describing the intent of
detection algorithms and supports the generator in organizing the source code.
As a case study, we apply \FOO to the generation of an intrusion detection
system for wireless sensor nodes. A side-by-side comparison shows that
\FOO-generated source code reduces message passing overhead, execution time and
memory footprint in comparison to sequential calls into standalone
implementations of the algorithms.

@@

The Internet of Things (IoT), the mother of all ubiquitous and pervasive
networks, populates our daily lives with millions of internet-connected
\emph{smart things}: from smart watches that assist visually impaired
persons\cite{porzi2013smart} to monitoring systems for waste management in our
\emph{smart cities}\cite{zanella2014internet}. Their applications cannot be
more diverse, however they share one common concern: each of these
\emph{things} is resource-constrained, relying on a battery and operating with
limited processing power and memory.

@@

Monitoring human activities is an active research field for more than a
decade\cite{lara2013survey}, but with the advent of the IoT it now moves into
the real world at an increased pace. Equipped with wearables, hosting numerous
sensors, the possibilities seem endless, but so are the number of detection and
recognition algorithms that have to run simultaneously on these small
\emph{things}, to support our augmented reality.

Another prime example of a class of algorithms that inhibit the same properties
is Intrusion Detection (ID). The large number of security related opportunities
for attackers requires an equally sized amount of ID algorithms, all with a
similar structure, still independent in their operation.

@@

The contributions of this work are threefold: the first contribution consists
in the formulation of a design pattern for detection/recognition algorithms for
IoT devices. This defines the target algorithms that fit our proposed
framework. The second contribution is a framework called \FOO.\@ It enables the
formal description of algorithms on a functional level and provides a code
generator producing source code for a given platform and configuration. This
way, \FOO also addresses the heterogeneous nature of both IoT and the
algorithms. The third contribution introduces functional code fusion (FCF) as a
source code generation paradigm to address the combination of algorithms. FCF
identifies common data and functions, and organizes code to eliminate redundant
iterations, tests and computations. Results show reduced usage of the wireless
radio, execution time and memory usage.

@@

  \item[Reimplementing] algorithms manually to eliminate these issues is
  not a one-time cost. A release of a new version of an algorithm, or simply a
  reselection of algorithms, due to a changing requirements, can trigger it.
  Also, changing, or simply implementing the algorithms from scratch, holds the
  risk of making mistakes.

@@

\begin{algorithm}[t]
  \small
  \caption{Detection/Recognition Algorithm Pattern}
  \label{alg:pattern}
  \begin{algorithmic}
    \Require{readings: global variables to hold collected sensor data for the given algorithm}
    \Require{devices: global variable to information about neighbor devices} \Comment{optional in case of distributed algorithms}
    \Function{collect\_data}{}
      \ForEach{$sensor \in required\_sensors$}
        \Let{values[]}{$sensor$}   \Comment{collect sensor data}
        \Let{$reading_{sensor}$}{\Call{interpret}{values[]}}  \Comment{update interpreted sensor info}
      \EndFor
      \State \Call{send}{$devices_x$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
    \EndFunction
    \Function{do\_housekeeping}{}
      \ForEach{$reading \in readings$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
        \If{$reading > \dots$} \Comment{validate recorded value}
          \State \dots \Comment{take actions}
          \State \Call{send}{$device_x$, ``info''} \Comment{optionally communicate} \label{alg:id-algo-pattern-send2}
        \EndIf
      \EndFor
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

The introduction of a code generator in the development process changes the way
of working: the source code is no longer the driving force, controlled by
developers. It is literally an intermediate representation between code
generator and compiler. A code generator constructs this representation using
three components, as shown in figure \ref{fig:design}: 1) the description of
the algorithm in the DSL, 2) information about the functional domains, and 3)
information about the platform and target language. At the level of the
compiler, a software library offloads specialized implementations and allows
custom extensions. The applied grayscale indicates the components' platform
dependency. This hierarchy is important from a development effort point of
view. The level of dependency on the target platform dictates the level of
effort to implement changes. This is important in relation to the scalability
of the framework itself.

@@

It's important to remember that a DSL is merely a \emph{thin veneer over a
(semantic) model} \cite{fowler2010domain}. Its sole purpose should be
populating the model. \FOO's DSL and model focus on avoiding constructions that
result in problems concerning loops, common data and the use of the network. To
do this, \FOO uses an event-based approach to defining functionality.
Describing algorithms based on events removes the technical interpretations and
allows FCF to identify the intent of algorithms and avoid the identified source
code problems.

@@

Different detection algorithms typically deal with the same classes of
information, be it sensor data or network messages. Functional domains
consolidate these and store and manage this data in a unified way. \FOO
provides the concept of such domains as plug-ins to the DSL, extending it with
domain specific storage and logic. For example in the case of distributed
algorithms a \emph{Nodes} domain will manage information about neighbor nodes,
exposing them as global objects to the algorithms, relieving them from managing
these themselves. Functional domains are also accessible as collections,
allowing execution of functions in their scope, again eliminating the need for
explicit loops.

@@

To convert the DSL-based ID algorithms into production-ready code, \FOO applies
code generation techniques. Although different approaches are applicable, like
template-based text expansion, XSLT transformations or CodeDom
\cite{dollard2004code}, we believe that the required transformations demand a
richer model. A semantic model \cite{fowler2010domain}, tied closely to the DSL
itself, can fulfill this role. Semantic transformations evolve the model into a
more syntax-oriented model, comparable to the idea of CodeDom. Syntactic
transformations can further evolve the model into source code. Figure
\ref{fig:code-generation} shows this process.

@@

\subsubsection{Functional code fusion (FCF)}

@@

A software library allows offloading of common implementation aspects, thus
avoiding small discrepancies in code and suboptimal use of shared resources.
Previously we illustrated this approach with a case for unified messaging:
parsing of incoming messages and grouping of outgoing messages. More common
functionality isn't generated verbatim either. A standard software library
provides functionality for the execution of functions based on a schedule or
the implementation of higher level conceptual functions. Optimization of these
type of algorithms is not within the scope of \FOO.\@ This further gives
freedom of choice to the users of the framework to select appropriate
implementations for this standard functionality.

@@

We also instrumented the code and evaluated four criteria: network usage
(number of packets and bytes), the time required to perform once cycle of the
event loop and the resulting image size. We collected these metrics for four
situations: without ID algorithms, with a watchdog, with reputation-tracking
and with both algorithms. The manual implementation realized both algorithms as
standalone modules and sequentially called them from the base-application's
event loop. Given the \FOO descriptions of the algorithms, the code generator
generated all four cases. Table \ref{tbl:results} shows the collected data.

@@

Many IoT applications typically are constructed of similar algorithms that
adhere to a pattern of collecting sensor data and periodically checking
recorded values for patterns that require action. Fusing these algorithms is
hard but needed to deal with the resource-constrained devices they are
targeting. Frequent reselection, e.g. due to changing requirements or changing
security threats, requires a flexible way to incorporate these algorithms in
subsequent iterations of the software.

@@

Initial experiments confirm the theoretically positive impact on resource
usage: grouping of messages lowers the number of packets by 13\%, while the
total number of bytes sent over the network is reduced by 7\%. The execution
time is improved by 7\% and memory usage can be reduced even by 20\%.

>>>>

The ability to detect patterns in its application context is what makes
\emph{things} \emph{smart}. The more patterns it can detect, the more accurate
its operational results will be. However, supporting an adequate set of pattern
detection algorithms imposes significant overhead on resource-constrained
devices such as those in the Internet of Things. Manual fusion of the
algorithms reduces resource consumption by optimizing resource sharing, yet,
proves to be time-consuming, repetitive and error-prone. To address this
problem we propose \FOO, a framework for the generation of efficient
combinations of detection algorithms. It consists of a domain specific language
and a code generator. The language allows formally describing the intent of
detection algorithms and supports the generator in organizing the source code.
As a case study, we apply \FOO to the generation of an intrusion detection
system for wireless sensor nodes. A side-by-side comparison shows that
\FOO-generated source code reduces message passing overhead, execution time and
memory footprint in comparison to sequential calls into standalone
implementations of the algorithms.

@@

The Internet of Things (IoT), the mother of all ubiquitous and pervasive
networks, populates our daily lives with millions of internet-connected
\emph{smart things}: from smart watches that assist visually impaired
persons\cite{porzi2013smart} to monitoring systems for waste management in
\emph{smart cities}\cite{zanella2014internet}. Their applications cannot be
more diverse, however they share one common concern: each of these
\emph{things} is resource-constrained, relying on a battery and operating with
limited processing power and memory.

@@

Monitoring human activities is an active research field for more than a
decade\cite{lara2013survey}, but with the advent of the IoT it now moves into
the real world at an increased pace. Equipped with wearables, hosting numerous
sensors, the possibilities seem endless, but so are the number of detection and
recognition algorithms that have to run simultaneously on these small
\emph{things}.

Another prime example of a class of algorithms that inhibit the same properties
is intrusion detection (ID). The large number of security related opportunities
for attackers requires an equally sized amount of ID algorithms, all with a
similar structure, still independent in their operation.

@@

The contributions of this work are threefold: the first contribution consists
in the formulation of a design pattern for detection/recognition algorithms for
IoT devices. This defines the target algorithms that fit our proposed
framework. The second contribution is a framework called \FOO.\@ It enables the
formal description of algorithms on a functional level and provides a code
generator producing source code for a given platform and configuration. This
way, \FOO also addresses the heterogeneous nature of both the IoT and the
algorithms. The third contribution introduces functional code fusion (FCF) as a
source code generation paradigm to address the combination of algorithms. FCF
identifies common data and functions, and organizes code to eliminate redundant
iterations, tests and computations. Results show reduced usage of the wireless
radio, execution time and memory usage.

@@

  \item[Reimplementing] algorithms manually to eliminate these issues is not a
  one-time cost. A release of a new version of an algorithm, or simply a
  reselection of algorithms, due to changing requirements, can trigger it.
  Also, changing, or simply implementing the algorithms from scratch, holds the
  risk of making mistakes.

@@

\begin{algorithm}[t]
  \small
  \caption{Detection/recognition algorithm pattern}
  \label{alg:pattern}
  \begin{algorithmic}
    \Require{readings: global variable to hold collected sensor data for the given algorithm}
    \Require{devices: global variable to store information about neighbor devices} \Comment{optional in case of distributed algorithms}
    \Function{collect\_data}{}
      \ForEach{$sensor \in required\_sensors$}
        \Let{values[]}{$sensor$}   \Comment{collect sensor data}
        \Let{$reading_{sensor}$}{\Call{interpret}{values[]}}  \Comment{update interpreted sensor info}
      \EndFor
      \State \Call{send}{$devices_x$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
    \EndFunction
    \Function{do\_housekeeping}{}
      \ForEach{$reading \in readings$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
        \If{$reading > \dots$} \Comment{validate recorded value}
          \State \dots \Comment{take actions}
          \State \Call{send}{$device_x$, ``info''} \Comment{optionally communicate} \label{alg:id-algo-pattern-send2}
        \EndIf
      \EndFor
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

The introduction of a code generator in the development process changes the way
of working: the source code is no longer the driving force, controlled by
developers. It is literally an intermediate representation between code
generator and compiler. A code generator constructs this representation using
three components, as shown in figure \ref{fig:design}: 1) the description of
the algorithm in the DSL, 2) information about the functional domains, and 3)
information about the platform and target language. At the level of the
compiler, a software library offloads specialized implementations and allows
for custom extensions. The applied grayscale indicates the components' platform
dependency. This hierarchy is important from a development effort point of
view. The level of dependency on the target platform dictates the level of
effort to implement changes. This is important in relation to the scalability
of the framework itself.

@@

It's important to remember that a DSL is merely a \emph{thin veneer over a
(semantic) model} \cite{fowler2010domain}. Its sole purpose should be
populating the model. \FOO's DSL and model focus on avoiding constructions that
result in problems concerning loops, common data and the use of the network. To
do this, \FOO uses an event-based approach to define functionality. Describing
algorithms based on events removes the technical interpretations and allows FCF
to identify the intent of algorithms and avoid the identified source code
problems.

@@

Different detection algorithms typically deal with the same classes of
information, be it sensor data or network messages. Functional domains
consolidate these and store and manage this data in a unified way. \FOO
provides the concept of such domains as plug-ins to the DSL, extending it with
domain specific storage and logic. For example in the case of distributed
algorithms a \emph{Nodes} domain will manage information about neighbor nodes,
exposing them as objects to the algorithms, relieving them from managing these
themselves. Functional domains are also accessible as collections, allowing
execution of functions in their scope, again eliminating the need for explicit
loops.

@@

To convert the DSL-based algorithms into production-ready code, \FOO applies
code generation techniques. Although different approaches are applicable, like
template-based text expansion, XSLT transformations or CodeDom
\cite{dollard2004code}, we believe that the required transformations demand a
richer model. A semantic model \cite{fowler2010domain}, tied closely to the DSL
itself, can fulfill this role. Semantic transformations evolve the model into a
more syntax-oriented model, comparable to the idea of CodeDom. Syntactic
transformations can further evolve the model into source code. Figure
\ref{fig:code-generation} shows this process.

@@

\subsubsection{Functional code fusion}

@@

A software library allows offloading of common implementation aspects, thus
avoiding small discrepancies in code and suboptimal use of shared resources.
Previously we illustrated this approach with a case for unified messaging:
parsing of incoming messages and grouping of outgoing messages. More common
functionality isn't generated verbatim either. A standard software library
provides functionality for the execution of functions based on a schedule or
the implementation of higher level conceptual functions. Optimization of this
type of algorithms is not within the scope of \FOO.\@ This further gives
freedom of choice to the users of the framework to select appropriate
implementations for this standard functionality.

@@

We also instrumented the code and evaluated four criteria: network usage
(number of packets and bytes), the time required to perform one cycle of the
event loop and the resulting image size. We collected these metrics for four
situations: without ID algorithms, with a watchdog, with reputation-tracking
and with both algorithms. The manual implementation realized both algorithms as
standalone modules and sequentially called them from the base-application's
event loop. Given the \FOO descriptions of the algorithms, the code generator
generated all four cases. Table \ref{tbl:results} shows the collected data.

@@

IoT applications are typically constructed of similar algorithms that adhere to
a pattern of collecting sensor data and periodically checking recorded values
for patterns that require action. Fusing these algorithms is hard but needed to
deal with the resource-constrained devices they are targeting. Frequent
reselection, e.g. due to changing requirements or changing security threats,
requires a flexible way to incorporate these algorithms in subsequent
iterations of the software.

@@

Initial experiments confirm the expected positive impact on resource usage:
grouping of messages lowers the number of packets by 13\%, while the total
number of bytes sent over the network is reduced by 7\%. The execution time is
improved by 7\% and memory usage can be reduced by 20\%.
